Cover letter for the original IJRR submission

### 2017-12-13 review_tro_2/reviewer_response.tex

First of all, we would like to thank all the Reviewers and the associate editor (AE) for considering our resubmission.
We are glad to find them satisfied with our modifications, and made our best to answer to the remaining comments as well as possible.
In the following lines, we will quote all the remaining comments and answer to explain how exactly we updated the paper to address them.

\section{Answers to the associate editor}

\quot {The C++/Java-style code listings are unnecessary and do not conform
to journal standards.  The authors rewrite their pseudocode to use
rigorous mathematical notation.  Suggest replacing "object.attribute"
with "Attribute(object)", and describing the attributes of the objects
in the text.
}

\as{We rewrote Algorithm 1, 2 and 3 accordingly to the recommendations of the Associate Editor, which we thank for the comment that allowed us to spot
a typo in pseudocode. Furthermore, it also allowed us to reduce the length of the pseudocode.}
\leavevmode 
\quot {The URLs in the paper seem rather brittle and are not suitable for
archival purposes.  Two URLs point to specific branches of various
Github repositories, and another refers to the documentation of a
Blender tool.  The authors should create a static web page for their
software which is guaranteed to persist for years to come... }

\as{Thanks for the comment. In fact, the github urls are not pointing to branches, but to release versions of the code, presented as zip files, which are in fact permanent
links. However, we agree with the Associate Editor that regrouping all the data and information on a single web page is relevant. So, we uploaded the source code to the first author
website (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html}), and provided all the required information to install the material in it. The user has the choice to download the code either from the github permanent link or directly from the author's website directly.}

\leavevmode 
\quot {... and to refer
to the underlying method used by the Blender tool.}

\as{Regarding the decimate tool from Blender, we proceeded as follows: in the paper, we mention a SIGGRAPH paper as one option to reduce mesh complexity, and mention that equivalent methods are proposed by many 3D authoring tools, such as Blender which we used. Then, in the static web page, we remind this piece of information and provide a permanent link to the version of Blender we used, along with the documentation that explains how to use the decimate tool.}

\qt{The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method[37].
Variations of this method are commonly implemented in most authoring tools. For our experiments we used the Blender decimate tool. Details of its use can be found in the static webpage associated with this paper \url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#decimate}.\\}

@@

\leavevmode \\
\section{Answers to Reviewer 2}
\quot{All the issues I raised in the first review have been addressed in
the revised version. I'm satisfied with most of the answers and the
changes made to the paper. In particular the novelty of the paper with
respect to the conference works is clarified. I appreciate also the new
structure of the manuscript that integrates material previously
relegated to the Appendices into the main body. Now, this work seems to
be self contained enough for a journal. }
\as{Thanks for the comment. We agree with the reviewer that the paper is better this way.}

\leavevmode \\
\quot{
1. I understand the point of view of the authors who focus on
multi-contact planning and would like to refer only briefly to previous
work in motion planning for multi-legged, statically-stable robots. In
principle that is OK, but I suggest to extend the paragraph added at
the end of Section I.A by few sentences that clarify what are the
advantages of your motion planning system with respect to these older
approaches. You are focusing on humanoids, for which these gains are
pretty obvious, but what about multi-legged robots? As an inspiration I
recommend this youtube video showing a simulated hexapod that does
something similar to your HyQ simulated example
\url{https://www.youtube.com/watch?v=Xio1TbjAK80}}

\as{TODO}
\leavevmode \\

\quot{
While the answer about the environment model is satisfying, I think
that you should no only refer to the FCL library in a footnote, but to
cite the paper that explains how it works.
}
\as{Thanks for the suggestion. The citation has indeed been included, as reference [31]TODO}
\leavevmode \\

### 2017-12-13 body/appendix_optim.tex
final changes from rev 1
webpage and links

% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.

@@

where $\dot{L}$ is the angular momentum expressed at the com.
Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:

>>>>

% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#optim} ), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.

@@

\noindent where $\dot{L}$ is the angular momentum expressed at the com.
Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:

### 2017-12-13 body/appendix_roms.tex
webpage and links

To generate a volume $W^k$, we proceed as follows:
\begin{enumerate}
\item Generate randomly $N$ valid limb configurations for $R^k$, for $N$ really large (say $100000$);
\item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
\item The resulting polytope can contain a very large number of faces. A last step is thus to simplify it in a conservative way with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}). For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
\end{enumerate}

>>>>

To generate a volume $W^k$, we proceed as follows:
\begin{enumerate}
\item Generate randomly $N$ valid limb configurations for $R^k$, for $N$ really large (say $100000$);
\item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
\item The resulting polytope can contain a very large number of faces.  A last step is thus to simplify it using an incremental decimation method~\cite{Garland:1997:SSU:258734.258849}.
Variations of this method are commonly implemented in most authoring tools. For our experiments we used the blender decimate tool. Details of its use can be found in the static webpage associated with this paper (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#decimate}). For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
%~ (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}). For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
\end{enumerate}

### 2017-12-13 body/results.tex
r2 typos
webpage and links

\subsection{Experimental validation of the contact plans} \label{sec:valid}
%~ \adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
The resulting movements have been validated either on the real HRP-2 robot (details can be found in~\cite{Carpentier2016}), or with our dynamic simulator, based on a state-of-the-art algorithm~\cite{Kaufman2008}.
In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}. The code source of the simulator is available at \url{https://github.com/andreadelprete/pinocchio_inv_dyn/releases/tag/rbprm}.
This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
The companion video shows the obtained motions.

@@

\begin{table*}[b!t!p!]
\centering
\footnotesize
\begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
  Scenario (nb steps) &  Complete guide generation (ms) & Static equilibrium (ms) & Collision (ms) & Inverse Kinematics (ms) & Total generation time (ms) & Time per step (ms)\\
 \hline
   Stairs (18) 	& 5 -- \textbf{6} --  18 		 & 13 --  \textbf{32} -- 329   	& 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
   Standing (24)& 65 -- \textbf{1086} --  5227   & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
   Car (86)& 320 -- \textbf{6971} --  44002 & 409 --  \textbf{1766} -- 14752   	& 297 -- \textbf{1187} -- 8483 & 3154 --  \textbf{15323} -- 165541 & 5834 --  \textbf{31391} -- 281000 & \textbf{365}\\
   Rubble (82)& 37 -- \textbf{573} --  1685 & 583 --  \textbf{2714} -- 9459 & 491 --  \textbf{1971} -- 6273 & 269 --  \textbf{706} -- 3118 & 1811 --  \textbf{7195} -- 23241 & \textbf{86} \\
   Race (134)& 14 -- \textbf{51} --  125 & 455 --  \textbf{1359} -- 21045   & 397 --  \textbf{923} -- 9924 & 228 --  \textbf{471} -- 5415 & 1436 --  \textbf{3343} -- 41446 & \textbf{25}
 \end{tabular}
\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process (not all parts are timed,
thus the average total computation time is higher than the sum of each part). The last
column indicates the average time necessary to compute one contact transition. The Collision column times includes the (negligeable) octree intersection operation necessary to retrieve the candidate samples.}
\label{tab:requestime}
\quad
\end{table*}

>>>>

\subsection{Experimental validation of the contact plans} \label{sec:valid}
%~ \adnote{I added the description of the simulations here, but I wonder whether this part should not be moved to the beginning of the Results section.}
To generate continuous movements from our contact plans we used either the framework proposed in \cite{Carpentier2016}, or our own implementation of a $\mathcal{P}_3$ solver (Appendix~\ref{app:optim}).
The resulting movements have been validated either on the real HRP-2 robot (details can be found in~\cite{Carpentier2016}), or with our dynamic simulator, based on a state-of-the-art algorithm~\cite{Kaufman2008}.
In the simulations we controlled the robot with a standard inverse-dynamics controller~\cite{DelPrete2015b}. The code source of the simulator is available at \url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#simulator}.
This controller tries to follow the given whole-body trajectories, giving higher priority to the center-of-mass and end-effectors tracking with respect to the joint tracking.
The controller also makes sure that the resulting contact forces lie inside the specified friction cones (we used a friction coefficient of 0.3), and that the joint position, velocity and torque limits are satisfied.
The companion video shows the obtained motions.

@@

\begin{table*}[b!t!p!]
\centering
\footnotesize
\begin{tabular}{ >{\centering\arraybackslash}m{37pt} | >{\centering\arraybackslash}m{57pt} | >{\centering\arraybackslash}m{65pt} | >{\centering\arraybackslash}m{70pt} | >{\centering\arraybackslash}m{73pt} | >{\centering\arraybackslash}m{80pt} | >{\centering\arraybackslash}m{10pt}}
  Scenario (nb steps) &  Complete guide generation (ms) & Static equilibrium (ms) & Collision (ms) & Inverse Kinematics (ms) & Total generation time (ms) & Time per step (ms)\\
 \hline
   Stairs (18) 	& 5 -- \textbf{6} --  18 		 & 13 --  \textbf{32} -- 329   	& 1 --  \textbf{4} -- 38 & 26 --  \textbf{127} -- 1345 & 92 --  \textbf{261} -- 2174 & \textbf{15} \\
   Standing (24)& 65 -- \textbf{1086} --  5227   & 27 --  \textbf{144} -- 338   & 2 --  \textbf{12} -- 37 & 144 --  \textbf{1046} -- 2374 & 371 --  \textbf{2257} -- 7671 & \textbf{94}  \\
   Car (86)& 320 -- \textbf{6971} --  44002 & 409 --  \textbf{1766} -- 14752   	& 297 -- \textbf{1187} -- 8483 & 3154 --  \textbf{15323} -- 165541 & 5834 --  \textbf{31391} -- 281000 & \textbf{365}\\
   Rubble (82)& 37 -- \textbf{573} --  1685 & 583 --  \textbf{2714} -- 9459 & 491 --  \textbf{1971} -- 6273 & 269 --  \textbf{706} -- 3118 & 1811 --  \textbf{7195} -- 23241 & \textbf{86} \\
   Race (134)& 14 -- \textbf{51} --  125 & 455 --  \textbf{1359} -- 21045   & 397 --  \textbf{923} -- 9924 & 228 --  \textbf{471} -- 5415 & 1436 --  \textbf{3343} -- 41446 & \textbf{25}
 \end{tabular}
\caption{minimum, \textbf{average} and worst time (in ms) spent in the generation process for each scenario and each critical part of the generation process (not all parts are timed,
thus the average total computation time is higher than the sum of each part). The last
column indicates the average time necessary to compute one contact transition. The Collision column times includes the (negligible) octree intersection operation necessary to retrieve the candidate samples.}
\label{tab:requestime}
\quad
\end{table*}
